---
title: "Explore EMP Water Quality Data"
author: "Dave Bosworth"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float:
      collapsed: false
editor_options: 
  chunk_output_type: console
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs"),
      envir = globalenv()
    )
    })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

Explore the water quality data collected by EMP in order to process it correctly for inclusion in the drought environmental predictors data set. 

# Global code and functions

```{r load packages, message = FALSE, warning = FALSE}
# Load packages
library(tidyverse)
library(scales)
library(viridisLite)
library(sf)
library(mapview)
# Make sure we are using `deltamapr` version 1.0.0, commit 5b11044a14aada45562d92899e2d8db42c0e8daf
# install.packages("devtools") 
# devtools::install_github("InteragencyEcologicalProgram/deltamapr", ref = "5b11044a14aada45562d92899e2d8db42c0e8daf")
library(deltamapr)
library(contentid)
library(here)
library(conflicted)

# Declare package conflict preferences
conflicts_prefer(dplyr::filter())

# Check if we are in the correct working directory
i_am("drought_variables/explore_emp_wq_data.Rmd")

# Set default options for mapview - turn off basemap color shuffle
mapviewOptions(basemaps.color.shuffle = FALSE)
```

```{r global functions}
# Function to create custom individual mapview objects for a set of coordinates
mapview_from_coord <- function(sf, layer_name, pt_color) {
  mapview(
    sf,
    layer.name = layer_name,
    col.regions = pt_color,
    alpha.regions = 0.8,
    cex = 5,
    label = "Station"
  )
}

# Function to create heatmaps of sampling effort
heatmap_sampeff <- function(df) {
  df %>%
    mutate(Station = fct_rev(factor(Station))) %>%
    ggplot(aes(x = Year, y = Station, fill = num_samples)) +
    geom_tile() +
    scale_x_continuous(
      limits = c(1974, 2022),
      breaks = breaks_pretty(20), 
      expand = expansion()
    ) +
    scale_fill_viridis_c(name = "Number of Samples") +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      legend.position = "top"
    )
}
```


# Import data

Import EMP WQ data and station coordinates from the [EDI data package 458.10](https://portal.edirepository.org/nis/mapbrowse?packageid=edi.458.10).

```{r import emp wq data, message = FALSE}
# Register a contentid for the EMP WQ data and station coordinates from the EDI data package 458.10
# This only needs to be done once
# EMP station coordinates
# register(
#   "https://portal.edirepository.org/nis/dataviewer?packageid=edi.458.10&entityid=86dd696bc3f8407ff52954094e1e9dcf"
# )

# EMP WQ data:
# register(
#   "https://portal.edirepository.org/nis/dataviewer?packageid=edi.458.10&entityid=cf231071093ac2861893793517db26f3"
# )

# Define contentid for EMP station coordinates
id_emp_coord <- "hash://sha256/120d7790b505f22a103ee392d13ca1c02b37fb9842d549942580b04c94eb2c67"

# Define contentid for EMP WQ data
id_emp_wq <- "hash://sha256/4c4b56d7d2a888b35efe2a6f2818f2b8b2fab26d0adb8f1c73ef522996b26ea8"

# Resolve the contentid's storing local copies for faster import
file_emp_coord <- resolve(id_emp_coord, store = TRUE)
file_emp_wq <- resolve(id_emp_wq, store = TRUE)

# Import EMP WQ data and station coordinates from the local copies using their contentid's
df_emp_coord <- read_csv(file_emp_coord)

df_emp_wq <- read_csv(
  file = file_emp_wq,
  col_types = cols_only(
    Station = "c",
    Date = "c",
    Secchi = "d",
    SpCndSurface = "d",
    SpCndBottom = "d",
    WTSurface = "d",
    WTBottom = "d",
    DissAmmonia_Sign = "c",
    DissAmmonia = "d",
    DissNitrateNitrite_Sign = "c",
    DissNitrateNitrite = "d",
    DissOrthophos_Sign = "c",
    DissOrthophos = "d"
  )
)
```

Import spatial data including the station coordinates for biological data and the Delta subregion polygons from the Drought Synthesis.

```{r import spatial data, message = FALSE}
# Import station coordinates for biological data
df_bio_coord <- read_csv(here("Station_coordinates.csv"))

# Load Delta subregion polygons from the Drought Synthesis
sf_delta <- R_EDSM_Subregions_Mahardja_FLOAT %>% 
  select(SubRegion) %>% 
  # Remove a few subregions outside our area of interest
  filter(!SubRegion %in% c("San Francisco Bay", "South Bay", "Upper Napa River")) 
```


# Prepare Data

Spatial data for maps:

```{r prepare spatial data}
# Prepare station coordinates of biological data for map - separate by taxon so
  # they can be added as independent layers
ndf_bio_coord <- df_bio_coord %>%
  rename_with(str_to_title) %>% 
  drop_na(Lat_wgs84, Long_wgs84) %>%
  # Remove phytoplankton stations because we're using a different set of
    # environmental predictors for them
  filter(Taxon != "phytoplankton") %>% 
  st_as_sf(coords = c("Long_wgs84", "Lat_wgs84"), crs = 4326) %>% 
  # Convert CRS so that its the same as the region and subregion polygons
  st_transform(crs = st_crs(sf_delta)) %>% 
  nest(.by = Taxon, .key = "sf_coord") %>% 
  mutate(Taxon = str_to_title(Taxon)) %>% 
  arrange(Taxon)

# Prepare EMP station coordinates for map
sf_emp_coord <- df_emp_coord %>% 
  drop_na(Latitude, Longitude) %>% 
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>% 
  # Convert CRS so that its the same as the subregion polygons
  st_transform(crs = st_crs(sf_delta)) %>%
  replace_na(list(EndDate = "ongoing"))
```

EMP WQ data:

```{r prepare emp wq data}
# Prepare EMP WQ data for heatmap of sampling effort
df_emp_wq_c <- df_emp_wq %>% 
  mutate(
    Date = ymd(Date),
    # Add placeholder values for <RL records without RL values
    DissAmmonia = if_else(DissAmmonia_Sign == "<" & is.na(DissAmmonia), 0.01, DissAmmonia),
    DissNitrateNitrite = if_else(DissNitrateNitrite_Sign == "<" & is.na(DissNitrateNitrite), 0.01, DissNitrateNitrite),
    DissOrthophos = if_else(DissOrthophos_Sign == "<" & is.na(DissOrthophos), 0.01, DissOrthophos)
  ) %>% 
  # Remove records where values for all parameters are NA
  filter(!if_all(where(is.numeric), is.na)) %>% 
  # Only include 2021 and earlier
  mutate(
    Month = month(Date),
    Year = year(Date),
    .after = Date
  ) %>% 
  filter(Year <= 2021) %>% 
  # Drop "Sign" variables because they're no longer necessary
  select(!ends_with("_Sign"))
```


# Filter data spatially

In order to spatially filter the EMP WQ sampling locations, we need to find the Delta subregions where biological data has been collected and categorize them based on taxon. We'll then use the resulting subregion polygons to spatially filter the EMP WQ stations.

```{r categorize delta subregions}
# Filter and categorize polygons from sf_delta
sf_delta_bio <- ndf_bio_coord %>% 
  mutate(
    sf_subr = list(sf_delta),
    sf_subr_filt = map2(
      sf_subr, 
      sf_coord, 
      ~ st_join(.x, .y, join = st_intersects) %>% 
        drop_na(Station) %>% 
        distinct(SubRegion, geometry)
    )
  ) %>% 
  select(Taxon, sf_subr_filt) %>% 
  unnest(sf_subr_filt) %>% 
  arrange(Taxon) %>% 
  group_by(SubRegion) %>% 
  mutate(
    Taxon = str_flatten_comma(Taxon),
    Taxon = if_else(Taxon == "Benthic, Fishes, Zooplankton", "All Taxa", Taxon)
  ) %>% 
  ungroup() %>% 
  distinct() %>% 
  st_as_sf(crs = st_crs(sf_delta))

# Create list of two mapview layers of the subregion polygons (all and filtered)
  # to be used across maps
ls_subr_polygons <- list(
  mapview(
    sf_delta,
    col.regions = "grey",
    layer.name = "All Subregions",
    alpha.regions = 0.5,
    legend = FALSE
  ),
  mapview(
    sf_delta_bio, 
    zcol = "Taxon", 
    layer.name = "Filtered Subregions",
    label = "SubRegion"
  )
)
```

Let's visualize these subregions with the biological and EMP WQ stations overlayed.

## Subregions map with biological sampling locations

```{r map biological}
ndf_bio_coord %>% 
  mutate(
    point_color = plasma(n = 3),
    map_obj = pmap(list(sf_coord, Taxon, point_color), mapview_from_coord)
  ) %>% 
  pull(map_obj) %>% 
  append(ls_subr_polygons, after = 0) %>% 
  reduce(`+`)
```

## Subregions map with EMP WQ locations

```{r map emp wq}
sf_emp_coord %>% 
  mapview_from_coord("EMP WQ Stations", "yellow") %>% 
  append(ls_subr_polygons, after = 0) %>%
  reduce(`+`)
```

## Filter EMP WQ stations

After looking at the maps above, we'll add the "Franks Tract" subregion to the polygons used for spatially filtering the EMP WQ stations. Otherwise, the polygons used for filtering look good. We'll come back to how they are categorized later.

```{r spatially filter emp wq}
# Add Franks Tract subregion to the polygons used for filtering
sf_delta_bio_c <- sf_delta_bio %>% 
  bind_rows(sf_delta %>% filter(SubRegion == "Franks Tract")) %>% 
  replace_na(list(Taxon = "Extra"))

# Spatially join subregions to the EMP WQ coordinates and remove stations
  # outside of subregions
sf_emp_coord_sfilt <- sf_emp_coord %>% 
  st_join(sf_delta_bio_c, join = st_intersects) %>% 
  drop_na(SubRegion)

# Remove data for stations outside of subregions from EMP WQ dataset
df_emp_wq_sfilt <- df_emp_wq_c %>% filter(Station %in% unique(sf_emp_coord_sfilt$Station))
```


# Combine overlapping stations

In order to combine overlapping EMP WQ stations and to filter them temporally, let's take a look at a map of where the filtered EMP WQ stations are located and heatmaps of their sampling effort.

## Map of EMP WQ stations

```{r map emp wq sfilt}
sf_emp_coord_sfilt %>% 
  mapview_from_coord("EMP WQ Stations", "yellow") +
  mapview(
    sf_delta_bio_c, 
    zcol = "Taxon", 
    layer.name = "Filtered Subregions",
    label = "SubRegion"
  )
```

## Sampling Effort {.tabset .tabset-pills}

```{r create heatmap sampling effort emp}
# Create a tibble of parameters to be used for the heatmaps
df_emp_wq_param <- 
  tibble(
    Parameter = c(
      "WTSurface",
      "WTBottom",
      "SpCndSurface",
      "SpCndBottom",
      "Secchi",
      "DissAmmonia",
      "DissNitrateNitrite",
      "DissOrthophos"
    ),
    Pretty_Name = c(
      "Water Temperature (surface)",
      "Water Temperature (bottom)",
      "Specific Conductance (surface)",
      "Specific Conductance (bottom)",
      "Secchi depth",
      "Dissolved Ammonia",
      "Dissolved Nitrate + Nitrite",
      "Dissolved Ortho-phosphate"
    )
  )

# Create nested dataframe and generate heatmaps for each parameter
ndf_emp_wq <- df_emp_wq_param %>% 
  mutate(
    df_data = list(df_emp_wq_sfilt),
    df_data_filt = map2(df_data, Parameter, ~ drop_na(.x, all_of(.y)))
  ) %>% 
  add_row(
    Pretty_Name = "All Parameters", 
    df_data_filt = list(df_emp_wq_sfilt), 
    .before = 1
  ) %>% 
  mutate(
    plt_samp_effort = map(
      df_data_filt,
      ~ count(.x, Station, Year, name = "num_samples") %>%
        heatmap_sampeff()
    )
  )
```

```{r print heatmap sampling effort emp, warning = FALSE, echo = FALSE, results = "asis", fig.width = 8, fig.height = 7.5}
for (i in 1:nrow(ndf_emp_wq)) {
  # Create subheadings for each Parameter
  cat("### ", ndf_emp_wq$Pretty_Name[i], "\n\n")
  # Print plot
  print(ndf_emp_wq$plt_samp_effort[[i]])
  cat("\n\n")
}
```

## Combine Stations

Looking at the map of filtered EMP WQ stations and heatmaps of their sampling effort, we'll combine the following stations:

* D41 and D42
* MD10 and MD10A
* S42 and NZS42

```{r combine overlapping stations}
# Combine overlapping stations in both WQ and station coordinates dataframes
ls_emp_sfilt_c <- 
  lst(
    df_emp_wq_sfilt,
    sf_emp_coord_sfilt
  ) %>% 
  map(
    ~ mutate(
      .x,
      Station = case_when(
        Station %in% c("D41", "D42") ~ "D41-D42",
        Station %in% c("MD10", "MD10A") ~ "MD10-MD10A",
        Station %in% c("S42", "NZS42") ~ "S42-NZS42",
        .default = Station
      )
    )
  )
```


# Filter data temporally

Next, we'll filter stations temporally based on their time series completeness. We'll keep stations that have less than 10% missing data assuming a monthly sampling frequency. This will be for each station's overall sampling record - not for each individual parameter.

```{r temporally filter emp wq}
# Calculate total number of monthly sampling events expected between 1975-2021
yrs_range <- range(ls_emp_sfilt_c$df_emp_wq_sfilt$Year)
num_yrs_mo <- length(yrs_range[1]:yrs_range[2]) * 12

# Calculate threshold for number of sampling events (10% missing)
num_yrs_thresh <- round(num_yrs_mo - num_yrs_mo * 0.1)

# Determine which stations meet the time series completeness criteria
lt_sta_keep <- ls_emp_sfilt_c$df_emp_wq_sfilt %>% 
  distinct(Station, Month, Year) %>% 
  count(Station) %>% 
  filter(n >= num_yrs_thresh) %>% 
  pull(Station)

# Filter stations temporally in both WQ and station coordinates dataframes
ls_emp_sfilt_tf <- ls_emp_sfilt_c %>% map(~ filter(.x, Station %in% lt_sta_keep))
```

## Map of final EMP WQ stations

Let's take a look at a map of the final EMP WQ stations after spatial and temporal filtering. These will be used in the drought environmental predictors data set.

```{r map emp wq sfilt tfilt}
ls_emp_sfilt_tf$sf_emp_coord_sfilt %>% 
  mapview_from_coord("EMP WQ Stations", "yellow") +
  mapview(
    sf_delta_bio_c, 
    zcol = "Taxon", 
    layer.name = "Filtered Subregions",
    label = "SubRegion"
  )
```

## Sampling Effort {.tabset .tabset-pills}

Let's also take a look at heatmaps of monthly sampling effort for each parameter of the final EMP WQ stations after spatial and temporal filtering.

```{r create heatmap sampling effort emp final}
# Create nested dataframe and generate heatmaps for each parameter
ndf_emp_wq_f <- df_emp_wq_param %>% 
  mutate(
    df_data = list(ls_emp_sfilt_tf$df_emp_wq_sfilt),
    df_data_filt = map2(df_data, Parameter, ~ drop_na(.x, all_of(.y)))
  ) %>% 
  add_row(
    Pretty_Name = "All Parameters", 
    df_data_filt = list(ls_emp_sfilt_tf$df_emp_wq_sfilt), 
    .before = 1
  ) %>% 
  mutate(
    plt_samp_effort = map(
      df_data_filt,
      ~ distinct(.x, Station, Month, Year) %>% 
        count(Station, Year, name = "num_samples") %>%
        heatmap_sampeff()
    )
  )
```

```{r print heatmap sampling effort emp final, warning = FALSE, echo = FALSE, results = "asis", fig.width = 8}
for (i in 1:nrow(ndf_emp_wq_f)) {
  # Create subheadings for each Parameter
  cat("### ", ndf_emp_wq_f$Pretty_Name[i], "\n\n")
  # Print plot
  print(ndf_emp_wq_f$plt_samp_effort[[i]])
  cat("\n\n")
}
```


# Conclusions and Next Steps

* The fish sampling locations extend further west than the stations for benthic and zooplankton. We'll probably end up creating two versions of the EMP water quality data to be included in the environmental predictors data set - one for fish and another for benthic and zooplankton. For fish, we'll include all EMP WQ stations after spatial and temporal filtering. For benthic and zooplankton, we'll exclude stations D41-D42 and D6 which are outside of the spatial range for sampling for these two taxonomic groups.
* We'll include the following parameters in the environmental predictors data set: water temperature (surface), specific conductance or salinity (surface), Secchi depth, dissolved ammonia, dissolved nitrate + nitrite, and dissolved ortho-phosphate. Unfortunately, nutrient data wasn't collected at D10, D12, D16, and D22 from 1996-2016. We'll need to decide if we'll include these stations in the averages for the nutrient parameters. Also, EMP didn't start collecting dissolved ammonia until 1979.

